A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2021-10-02 analysis/paper/paper.Rmd
Both Dybdalshei sites appear fine in terms of rasters and 250 m distance

---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

# Introduction

# Background

# Methods

# Results

@@

# Discussion

# Conclusion

>>>>

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

# Introduction
The association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses both published in 1905 (see also @maccurdy1906). Ever since, this finding has been fundamental to Stone Age research in Norway. One of its major implications is that relative sea-level change can be used for the dating of the sites. This is based on the fact that since most Stone Age sites were situated on or close to the contemporary shoreline during their original occupation, a reconstruction of the trajectory of shoreline displacement can be used to date the sites based on their elevation relative to the present day sea level. The method of shoreline dating is still fundamental to most Stone Age research in Norway, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the generally coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most accurate method by which one can hope to date the sites. Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.

Despite the fundamental role of shoreline-dating in Norway, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement, it can potentially identify societal shifts if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the region.

# Background
After the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, meaning that the relative sea level has been dropping throughout prehistory. In other areas of Norway the situation is sometimes more complicated. In some areas relative sea-level change has transitioned between regression and transgression phases, and the archaeological remains can be destroyed. In other areas, the sea-level has been stable over longer periods of time, meaning coastal locations have often been reused multiple times over long time-spans, creating a palmpiset of settlement events that are impossible to disentangle. Comparatively, given the continuous and relatively rapid shoreline regression in south-eastern Norway, any one location in the region has only been shore-bound within a relatively limited time-span. As this general displacement trajectory is the same throughout the entirety of prehistory, this makes the region especially useful for testing of the assumption of a shore-bound settlement and to identify potential deviations from this pattern.

The most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves. If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this. In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of of the overlap between shoreline displacement curves and the radiocarbon probability distributions. Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensisble if the site was *not* shore bound [e.g. @jaksland2014]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, it would make intuitive sense that the location was used when the ocean reached the elevation of the ledge and the site would have been accessible by means of watercraft. This line of reasoning is often also used in survey projects which typically do not produce material suitable for radiocarbon dating. The evaluation of the degree to which sites were shorebound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations. Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evalation and quantification of this tendency has been done.

One of the only larger scale evaulations of this relationship in Norwegian archaeology has been done by Solheim and collegues (@solheim20, see also @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found a overlap of the confidence bands of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites. However, in several of the cases where there was a discrepancy, the

Given that the aim of this study is to test whether or not there is a match between dates provided by means of shoreline displaceent curves and radiocarbon dates, it is appropriate to evaulate this relationship by means of statistical equivalence tests. This effectively shifts the common direction of the null-hypothesis, giving H0: The date given by means of radiocarbon dating and by shoreline displacement are different. Ha: The date given by means of radiocarbon dating and by shoreline displacement are the same. While this consitutes the formal dimension of the study, there must necessarily also be a qualitative and more subjectivly driven dimension as well. As was exemplified in Solheims study, and which has been convincingly argued in several individual studies, many of the sites where there is a mismatch between the radiocarbon dates and the shoreline dates, the assemblages are clearly related to the shoreline date rather than the radiocarbon dates. Whether these mismatches represent later shorter visits responsible for the younger radiocarbon dates or whether they are the result of contaminated samples or by pedoturbation can be difficult to evaluate and will here be entirely dependent on the discretion of the original excavation report. While such evaluations can be difficult, it is nonetheless a necessity that will be held strictly separate from the formal equivalence tests in the analysis to follow.

# Methods
Relative sea level is so named because the parameters of interest is the height of the sea surface relative to its present level. Sea level is here understood as the distance from the solid surface of the Earth to the mean sea level as averaged over time, also known as the geoid. Variation in this distance is determined by effects that can be divided into three main categories. The first of these is eustatic sea-level change. Eustatic sea level is understood as the sea level if the water has been evenly distributed across the Earth's surface and without accounting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself. The eustatic sea-level is most notably impacted by glaciation and de-glaciation which can either bind or release large amounts of water into the oceans, as well as volumetric changes in the Earth's crust following from . The second category is that of istostasy, which pertains to adjustments in the Earth's crust that results from processes to regain gravitational equilibrium relative to the underlying viscous mantle. This is typically the result of erosion of the crust which causes its weight to be redistributed and the surface to be pushed upwards, or glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, which either causes the lithosphere to be pushed into the mantle or to rebound and to lift upwards. Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.

This study necessaitated the compilation of multiple relative sea-level curves developed for different parts of the region. Most of the curves come provided with a confidence interval which is largely based on an evaluation by the geologists who developed the curve. The reason why these confidence intervals are not strictly mathematically derived is that the curves do not only contain uncertainy as related to radiometric dates, which are fairly well understood, but also hold potential error as related to the interpretation of sediment cores, placement of samples,  direction and adjustment to a common isobase based on the direction and tilt of the variation in rebound intensity, as well as, to name but a few. The error bands are consequently complex, and cannot necessarily be directly compared between curves as these have been developed at different times, using a wide range of different methods, and often using older and outdated calibration curves for both marine and terrestrial samples. However, any reinterpretation and standardisation of the curves is also hampered by a range of issues. The original data or necessary details about the sampling processes are not always available, and a reinterpretation based soley on radiometric dates, without taking into account the considerations and evaluations of the field geologists can be entirely misleading. In a rudimentary attempt to account for these differences when compiling the curves, the width of the confidence band for the sea level at a single date across the entire region is given as the largerst uncertainty of any one of the curves at that point in time. This averages to an uncertainty of +/- calendar years. This uncertainty is uniformly distributed within the confidence band.

The goal is to test the synchroneity of two space-time processes, the occupation of a site and continuously shifting sea-levels. The synchroneity of these processes is tested by the comparison of singular events represented by the death of the dated organic materials. Additionally, defining what shore bound settlement represents can be problematic. In the literature this is often qualified with statements along the lines of "the sites were located on or close to the contemporenous shore-line". What constitutes close is of course highly variabl,e complicated further by the fact that vertical changes in sea level can results in widely different effects on the horisontal distance between a given point and the shore-line. Other elements of interest is the tidal range, and the exposure of the location in question, which can also have an impact on the distance of the site to the mean ocean surface.

@@
